Human capital mediates natural selection in contemporary humans

### 2022-01-10 why-natural-selection.Rmd
Tweaks for EHB submission

Living organisms evolve through natural selection, in which allele frequencies
change in the population through differential reproduction rates. Studying the
mechanisms behind natural selection can help us better understand how individual
differences in complex traits and disease risk arise [@benton2021influence].
Recent work confirms that natural selection is taking place in modern human
populations, using genome-wide analysis
[@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
particular, genetic variants associated with higher educational attainment are
being selected against, although effect sizes appear small.

@@

To correct for ascertainment bias, we weight participants using population
data. We try three alternative weighting schemes: (1) "Geographical" weighting,
by geography, age and presence/absence of a partner; (2) age and highest educational
qualification; and (3) for women only, age, highest qualification, and age at 
first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes, and also with uniform
weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
reaching significance under at least one scheme. Weighting makes
a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic variables.[^weighting]

[^weighting]: To make interpretation straightforward, we do not use weights in 
results below unless noted. However, all our results are robust to using weights.

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
predictions, along with corresponding empirical results for the correlation
between polygenic scores and fertility.[^one-period-results] The key insight of
the model is that for middling levels of risk aversion, the substitution effect
dominates at low income levels, but is balanced by the income effect as income
increases.

@@

Second, we run a mediation analysis to directly test whether the correlation
between each polygenic score and fertility is mediated by educational attainment
(Appendix Table \@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
indirect effect of the score on fertility *via* educational attainment is
significantly different from zero ($p < 0.05/33$). Among these scores, the
median proportion of the total effect explained by the indirect effect is 
`r percent(median_prop)`.

@@

These results support our theory that natural selection on polygenic scores is
driven by their correlation with human capital. We also test for causality by
running within-siblings regressions (Appendix Figure
\@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
within-sibling effects are insignificant after Bonferroni correction. However,
effect sizes are positively correlated with effect sizes from the pooled model,
and about 50% smaller (regressing within-sibling on pooled effect sizes, $b$ =
`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the ~50%
decrease in heritability in within-sibling GWASs on age at first birth and
educational attainment [@Howe2021.03.05.433935]. We see these results as
providing tentative evidence that polygenic scores cause fertility, with effects
being partly driven by gene-environment correlations. We also reran
within-siblings regressions adding a control for education. Most effect sizes
barely change, suggesting that our measure of education does not in general
mediate differences in fertility among siblings. However, for educational
attainment polygenic scores, effect sizes are reduced by `r percent(educ_reduction)`.

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, our effect sizes appear unchanged over a large
span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and
\@ref(tab:tbl-children-by-period)), during which government spending on
child-related benefits varied considerably [@hoc1999socsec]. In general, there
is only weak evidence that welfare benefits affect fertility
[@gauthier2007impact]. A second alternative theory is that polygenic scores
correlate with the motivation to have children, i.e. parameter $a$ in the model
[cf. @jones2008fertility]. This theory would not explain why selection effects
are smaller at higher incomes and education levels. (Indeed, in the model, the
effect of $a$ on fertility becomes stronger at higher levels of human capital.) A
third alternative is that traits under selection are linked to externalizing
behaviour and risk-seeking. This might be partially captured by our
risk-aversion parameter $\sigma$, which affects fertility by changing the
marginal utility of income, but a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced, or whether genetic
variants underlying other phenotypes will show a similar pattern to those
studied here. In addition, genetic effects on educational attainment have been
shown to be inflated in population-based samples as compared to within-family
designs, likely because of indirect genetic effects, gene-environment
correlations, and/or assortative mating [@lee2018gene; @selzam2019comparing;
@kong2018nature; @Howe2021.03.05.433935]. In short, it is probably too early to
tell whether modern natural selection has a substantively important effect on
population means of phenotypes under selection. Nevertheless, we note that selection
effects on our measured polygenic scores are still relatively small, even after
reweighting to account for ascertainment bias.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases the correlation of polygenic scores with
income for `r n_increases` out of 33 polygenic scores, with a median
proportionate increase of `r median_pct_change`% (Appendix Table
\@ref(tab:tbl-inequality-effects)).

>>>>

Living organisms evolve through natural selection, in which allele frequencies
change in the population through differential reproduction rates. Studying the
mechanisms behind natural selection can help us better understand how individual
differences in complex traits and disease risk arise [@benton2021influence].
Recent work confirms that natural selection is taking place in modern human
populations, using genome-wide analysis
[@Barban_2016;@beauchamp2016genetic;@conley2016assortative;@kong2017selection;@sanjak2018evidence;@FIEDER202216]. In
particular, genetic variants associated with higher educational attainment are
being selected against, although effect sizes appear small.

@@

To correct for ascertainment bias, we weight participants using population data.
We try three alternative weighting schemes: (1) "Geographical" weighting, by
geography, age and presence/absence of a partner; (2) age and highest
educational qualification; and (3) for women only, age, highest qualification,
and age at first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots
selection effects among the entire sample, estimated with the three weighting
schemes, and also unweighted. Effects for `r res_sig_all` out of 33 polygenic
scores are significant at $p < 0.05/33$ under all four weighting schemes, with
`r res_sig_any` scores reaching significance under at least one scheme.
Weighting makes a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic
variables.[^weighting]

[^weighting]: To make interpretation straightforward, we do not use weights in 
results below unless noted. However, all our qualitative results are robust to
using weights.

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
predictions, along with our corresponding empirical results for the correlation
between polygenic scores and fertility.[^one-period-results] The key insight of
the model is that for middling levels of risk aversion, the substitution effect
dominates at low income levels, but is balanced by the income effect as income
increases.

@@

Second, we run a mediation analysis to directly test whether the correlation
between each polygenic score and fertility is mediated by educational attainment
(Appendix Table \@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
indirect effect of the score on fertility *via* educational attainment is
significantly different from zero ($p < 0.05/33$). Among these scores, the
median proportion of the total effect explained by the indirect effect is 
`r percent(median_prop)`. The educational attainment variable is a relatively
crude measure of human capital: more accurate measures would likely
explain more of the total effect.

@@

These results support our theory that natural selection on polygenic scores is
driven by their *correlation* with human capital. We also test whether polygenic
scores *cause* fertility by running within-siblings regressions (Appendix Figure
\@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
within-sibling effects are insignificant after Bonferroni correction. However,
effect sizes are positively correlated with effect sizes from the pooled model,
and about 50% smaller (regressing within-sibling on pooled effect sizes, $b$ =
`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the ~50%
decrease in heritability in within-sibling GWASs on age at first birth and
educational attainment [@Howe2021.03.05.433935]. We see these results as
providing tentative evidence that polygenic scores cause fertility, with effects
being partly driven by correlations with environmental variation in human
capital. We also reran within-siblings regressions adding a control for
education. Most effect sizes barely change, suggesting that our measure of
education does not in general mediate differences in fertility among siblings.
However, for the educational attainment polygenic score (EA3), effect sizes are
reduced by `r percent(educ_reduction)`.

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, most effect sizes appear unchanged over a large
span of twentieth-century history (Appendix Tables
\@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)), during
which government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter
$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
selection effects are smaller at higher incomes and education levels. (Indeed,
in the model, the effect of $a$ on fertility becomes stronger at higher levels
of human capital.) A third alternative is that traits under selection are linked
to externalizing behaviour and risk-seeking. This might be partially captured by
our risk-aversion parameter $\sigma$, which affects fertility by changing the
marginal utility of income, but a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced, or whether genetic
variants underlying other phenotypes will show a similar pattern to those
studied here. In addition, effects of polygenic scores may be inflated in
population-based samples, because of indirect genetic effects, gene-environment
correlations, and/or assortative mating [@lee2018gene;
@selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935], although we do
not expect that this should impact their association with number of offspring,
or the resulting changes in allele frequencies. In short, it is probably too
early to tell whether modern natural selection has a substantively important
effect on population means of phenotypes under selection. Nevertheless, we note
that selection effects on our measured polygenic scores are still relatively
small, even after reweighting to account for ascertainment bias.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases the correlation of polygenic scores with
income for `r n_increases` out of 33 polygenic scores, with a median
percentage increase of `r median_pct_change`% (Appendix Table
\@ref(tab:tbl-inequality-effects)).
